---
title: "MC Model Comparisons"
author: "Group 3"
date: "2025-12-01"
output:
  pdf_document: default
  html_document: default
---
\colorlet{shadecolor}{gray!10}
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(MASS)
library(GGally)
library(tidymodels)
library(glmnet)
library(pROC)
library(randomForest)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE)
```
\newcommand{\hstart}{ \colorlet{shadecolor}{green!20}
\begin{shaded} }
\newcommand{\hstop}{  \end{shaded} \colorlet{shadecolor}{gray!10}}

\hstart

# Minimum Liklihood, Ridge, and Random Forest: A Monte Carlo Model Comparison

\hstop

\hstart

### 0. Source functions and simulated data for model comparisons

\hstop

```{r source functions}
# sourcing method functions

source("simulate_functions.R")  # for 

```

```{r simulated data}
# sourcing simulated data

simulated_data <- readRDS("data/simulated_data.rds")

```

\hstart

### 1. Maximum Liklihood

\hstop

```{r maximum likelihood}
# Maximum Likelihood


```


\hstart

### 2. Ridge Regression

\hstop

```{r ridge regression}
# ridge regression


```


\hstart

### 3. Random Forest

\hstop
```{r random forest}
# random forest


```


\hstart

### 4. Comparisons

\hstop

```{r comparisons}
# comparisons


```

