---
title: "MC - Ridge Regression"
author: "Andi Mellyn"
date: "2025-12-01"
output:
  html_document: default
  pdf_document: default
---

\colorlet{shadecolor}{gray!10}
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(MASS)
library(GGally)
library(tidymodels)
library(glmnet)
library(pROC)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE)
```
\newcommand{\hstart}{ \colorlet{shadecolor}{green!20}
\begin{shaded} }
\newcommand{\hstop}{  \end{shaded} \colorlet{shadecolor}{gray!10}}

\hstart

# Monte Carlo Simulation using a Ridge Regression Model

\hstop

```{r sourcing functions}
# sourcing the simulated functions

source("simulate_functions.R")

```

```{r ridge logistic regression function}
# ridge logistic regression function

fit_logistic_ridge <- function(dat) {
  # dat: data.frame with first column y and the rest predictors
  
  X <- as.matrix(dat[ , -1])  # all columns except the first
  y <- dat$y                  # first column is y
  
  # alpha = 0 -> ridge penalty
  cvfit <- cv.glmnet(
    x = X,
    y = y,
    alpha = 0, 
    family = "binomial"
  )
  
  cvfit
}

```

```{r evaluate ridge model}
# ridge-specific evaluate model

evaluate_ridge_model <- function(cvfit, dat) {
  X <- as.matrix(dat[ , -1])
  y <- dat$y
  
  # lambda.min from CV (the value that minimizes the error)
  probs <- predict(cvfit, newx = X, type = "response", s = "lambda.min")
  
  auc_val <- pROC::auc(y, probs)
  brier <- mean((y - probs)^2)
  
  tibble::tibble(
    AUC = as.numeric(auc_val), # AUC: area under the ROC curve
    Brier = brier, # brier: measure of the accuracy of probabilistic predictions
    lambda = cvfit$lambda.min
  )
}

```

```{r test run}
# testing the ridge model for one run

set.seed(400)

# simulating one dataset 
dat_test <- simulate_logit(N = 1000, P = 8, rho = 0.5, beta_pattern = "strong")

ridge_fit <- fit_logistic_ridge(dat_test)
ridge_summary <- evaluate_ridge_model(ridge_fit, dat_test)

ridge_summary

```

```{r Monte Carlo simulation}
# Monte Carlo simulation settings (subset of Van Calster et al. design)

EPV_vals <- c(5, 10, 50)    # subset -> research study uses 7 values
event_frac <- 0.5           # subset -> research study uses 4 fractions
P_vals <- c(4, 8, 12)       # same p-vals as study
rho_vals <- c(0, 0.5)       # same continuous predictors
beta_patterns <- c("equal", "strong", "noise", "halfnoise")   # same beta patterns

B <- 10  # number of Monte Carlo repetitions per condition


# Monte Carlo loop for ridge regression

set.seed(400)

ridge_results <- list()

iter <- 1

for (EPV in EPV_vals) {
  for (P in P_vals) {
    for (rho in rho_vals) {
      for (bp in beta_patterns) {
        for (b in 1:B) {
          
          # compute sample size N from EPV formula:
          # EPV = (N * event_frac) / P => N = EPV * P / event_frac
          N <- ceiling(EPV * P / event_frac)
        
          # simulate one dataset
          dat <- simulate_logit(N = N, P = P, rho = rho, beta_pattern = bp)
        
          # fit ridge model
          ridge_fit <- fit_logistic_ridge(dat)
        
          # evaluate performance on this dataset
          ridge_summary <- evaluate_ridge_model(ridge_fit, dat)
        
          # store results
          ridge_results[[iter]] <- tibble(
            method = "ridge",
            EPV = EPV,
            sim = b, 
            P = P, 
            rho = rho,
            beta_pattern = bp,
            AUC = ridge_summary$AUC,
            Brier = ridge_summary$Brier,
            lambda = ridge_summary$lambda
          )
          iter <- iter + 1
        }
      
      }
    
    }
  
  }
  
}


ridge_results_df <- bind_rows(ridge_results)
head(ridge_results_df)

```

```{r ridge summary stats}

# matching the research paper summary stats

ridge_condition_summary <- ridge_results_df |>
  group_by(EPV, P, rho, beta_pattern) |>
  summarise(
    n_sims = n(),
    mean_AUC = mean(AUC),
    sd_AUC = sd(AUC),
    mean_Brier = mean(Brier), 
    sd_Brier = sd(Brier),
    mean_lambda = mean(lambda),
    .groups = "drop"
  )

ridge_condition_summary
knitr::kable(ridge_condition_summary, digits = 3)

```

```{r AUC plot}
# AUC plot

ggplot(ridge_condition_summary,
       aes(x = factor(P), y = mean_AUC, group = factor(rho), color = factor(rho))) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ beta_pattern) +
  labs(
    title = "Ridge Regression: Mean AUC by Condition",
    x = "Number of Predictors (P)",
    y = "Mean AUC",
    color = "Correlation (rho)"
  ) +
  geom_ribbon(aes(ymin = mean_AUC - sd_AUC,
                ymax = mean_AUC + sd_AUC,
                fill = factor(rho)),
            alpha = 0.30, color = NA)

```

```{r Brier plot}
# Brier plot

ggplot(ridge_condition_summary,
       aes(x = factor(P), y = mean_Brier, group = factor(rho), color = factor(rho))) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ beta_pattern) +
  labs(
    title = "Ridge Regression: Mean Brier Score by Condition",
    x = "Number of Predictors (P)",
    y = "Mean Brier Score",
    color = "Correlation (rho)"
  ) +
  geom_ribbon(aes(ymin = mean_Brier - sd_Brier,
                ymax = mean_Brier + sd_Brier,
                fill = factor(rho)),
            alpha = 0.15, color = NA)
  


```

\hstart

### Calibration Plot
For illustration purposes we show a calibration plot for a SINGLE representative scenario:
$N = 1000, \\P = 8, \\rho = 0.5$, 
beta pattern = $"strong"$.

\hstop

```{r ridge calibration setup}
# calibration plot - how well do the predicted values line up with the actual observed values

set.seed(400)

# simulating a single dataset for calibration
dat_cal <- simulate_logit(
  N = 1000,
  P = 8, 
  rho = 0.5,
  beta_pattern = "strong"
)

# fit ridge logistic regression
ridge_fit_cal <- fit_logistic_ridge(dat_cal)

```

```{r ridge calibration data}

# design matrix for glmnet
X_cal <- as.matrix(dat_cal[ , -1])

# predicted probabilities from the ridge model
ridge_probs <- predict(
  ridge_fit_cal,
  newx = X_cal,
  type = "response", 
  s = "lambda.min"
)

# build calibration dataframe
ridge_cal_df <- dat_cal |>
  mutate(prob = as.numeric(ridge_probs),
         decile = ntile(prob, 10)) |>
  group_by(decile) |>
  summarise(obs = mean(y), pred = mean(prob), .groups = "drop")

ridge_cal_df
  
```

```{r calibration plot}

ggplot(ridge_cal_df, aes(x = pred, y = obs)) +
  geom_point(size = 3) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, color = "limegreen", linetype = 2) +
  xlim(0, 1) + ylim(0, 1) +
  labs(
    title = "Calibration Plot – Ridge Logistic Regression"
  )

```

\hstart

# Compare Ridge to MLE and Forest

\hstop

