---
title: "MC Model Comparisons"
author: "Group 3: Andi Mellyn, Jessica Reyes, Hope Winsor"
date: "2025-12-01"
output:
  pdf_document: default
  html_document: default
---
\colorlet{shadecolor}{gray!10}
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(MASS)
library(GGally)
library(tidymodels)
library(glmnet)
library(pROC)
library(randomForest)
library(dplyr)
library(here)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE)
```

\newcommand{\hstart}{ \colorlet{shadecolor}{green!20}
\begin{shaded} }
\newcommand{\hstop}{  \end{shaded} \colorlet{shadecolor}{gray!10}}

\newcommand{\pstart}{ \colorlet{shadecolor}{gray!20}
\begin{shaded} }
\newcommand{\pstop}{  \end{shaded} \colorlet{shadecolor}{gray!10}}

```{r}
# helper files so we don't rerun data simulations
mle_path   <- here("MLE_Simulation.csv")
ridge_path <- here("ridge_mc_testing_results.csv")
rf_path    <- here("RandomForest_Simulation.csv")

ridge_extra_path <- here("simulations_rmse_mae", "ridge_simulation_extra_values.csv")
mle_extra_path   <- here("simulations_rmse_mae", "mle_simulation_extra_values.csv")
rf_extra_path    <- here("simulations_rmse_mae", "random_forest_simulation_extra_values.csv")

```


\hstart

# Maximum Liklihood, Ridge, & Random Forest: A Monte Carlo Model Comparison

\hstop

\hstart

## Source functions and simulated data for model comparisons

\hstop

```{r source functions}
# simulate data -> logistic regression

simulate_logit <- function(N, P, rho, beta_pattern) {
  
  # Predictor distribution 
  # Correlation = rho (0 or 0.5)
  Sigma <- matrix(rho, P, P)
  diag(Sigma) <- 1
  
  X <- mvrnorm(n = N, mu = rep(0, P), Sigma = Sigma)
  
  # Predictor effects 
  if (beta_pattern == "equal") {
    beta <- rep(0.5, P)
    
  } else if (beta_pattern == "strong") {
    beta <- c(1.0, rep(0.2, P-1))      # 1 strong effect
    
  } else if (beta_pattern == "noise") {
    beta <- c(0, rep(0.3, P-1))        # first predictor is noise
    
  } else if (beta_pattern == "halfnoise") {
    beta <- c(rep(0, P/2), rep(0.3, P/2))  # first half noise
    
  } else {
    stop("Unknown beta pattern")
  }
  
  # Linear predictor + logistic link
  eta <- X %*% beta
  pi  <- 1/(1 + exp(-eta))
  
  # Binary outcome from Bernoulli
  y <- rbinom(N, size = 1, prob = pi)
  
  data.frame(y = y, X)
}

```

\hstart

## Background

\hstop

\pstart

Binary logistic regression is a common method for developing clinical prediction models. Traditionally, guidance on choosing an appropriate sample size has relied on the concept of events per variable (EPV), defined as the ratio of the number of outcome events to the number of degrees of freedom, think signal to noise ratio. The "best" EPV value has been $\ge 10$. In practice, EPV is used to determine both the minimum sample size required for a study and the maximum number of predictors that can be included in a model.

The study *Sample size for binary logistic prediction models: Beyond events per variable criteria* challenges the adequacy of this rule by examining whether EPV is truly the primary driver of out-of-sample predictive performance. Using a large Monte Carlo simulation study, van Smeden et al. evaluated predictive accuracy, calibration slope, AUC, and brier scores across 4,032 different combinations of data-generating scenarios run on 7 different models including maximum likelihood, backwards subset selection, firth's penalized likelihood, heuristic shrinkage, ridge, and LASSO. Their results show that EPV alone has a weak relationship with predictive performance.

For our project, we focused on a subset of original simulation design. We replicated the core comparison between maximum likelihood and ridge logistic regression to examine how shrinkage affects predictive performance under varying EPV conditions. In addition, we extended the original framework by incorporating a random forest classifier, allowing us to assess whether a non-parametric machine learning approach exhibits similar sensitivity to EPV and sample size. By comparing these three modeling strategies under a shared simulation structure, we aimed to evaluate whether the limitations of EPV observed in logistic regression also persist in more flexible predictive models.

\pstop

\hstart

## 1. Maximum Liklihood

\hstop

```{r maximum likelihood}
# Maximum Likelihood
# Data Simulation Function

data_simulation_function <- function(N, P, rho, beta_pattern) {
  Sigma <- matrix(rho, P, P)
  diag(Sigma) = 1
  
  X <- mvrnorm(n = N, mu = rep(0, P), Sigma = Sigma)
  
  # Set coefficients
  if (beta_pattern == "equal") {
    beta <- rep(0.5, P)
  } else if (beta_pattern == "strong") {
    beta <- c(1, rep(0.2, P-1))
  } else if (beta_pattern == "noise") {
    beta <- c(0, rep(0.3, P-1))
  } else if (beta_pattern == "halfnoise") {
    beta <- c(rep(0, P/2), rep(0.3, P/2))
  } else stop("Unknown beta pattern")
  
  eta <- X %*% beta
  p <- 1 / (1 + exp(-eta))
  
  y <- rbinom(N, 1, p)
  
  data.frame(y = y, X)
}

```

```{r}
# Model Fitting

model_fitting_function <- function(train_data) {
  glm(y ~ ., data = train_data, family = binomial)
}

```

```{r}
# Model Assessment function

model_assessment_function <- function(model, test_data) {
  probs <- predict(model, newdata = test_data, type = "response")
  
  auc_val <- auc(test_data$y, probs)
  brier <- mean((test_data$y - probs)^2)
  
  tibble(
    AUC = as.numeric(auc_val),
    Brier = brier
  )
}

```

```{r}

fit_logistic_mle <- function(dat) {
  glm(y ~ ., data = dat, family = binomial)
}

evaluate_mle_model <- function(fit, dat) {
  probs <- predict(fit, newdata = dat, type = "response")
  auc_val <- auc(dat$y, probs)
  brier <- mean((dat$y - probs)^2)
  
  tibble(AUC = as.numeric(auc_val), Brier = brier, lambda = NA)
}

```


```{r mle test run}
# testing the mle model for one run

set.seed(400)

# simulating one dataset 
dat_mle <- data_simulation_function(N = 1000, P = 8, rho = 0.5, beta_pattern = "strong")
fit_mle <- model_fitting_function(dat_mle)

mle_fit <- model_fitting_function(dat_mle)
mle_summary <- model_assessment_function(fit_mle, dat_mle)

mle_summary

```

```{r mle_mc, eval=FALSE}

set.seed(400)

# Simulation loop

EPV_vals <- c(5, 10, 50)
event_frac <- 0.5    

P_vals <- c(4, 8, 12)
rho_vals <- c(0, 0.5)
beta_patterns <- c("equal", "strong", "noise", "halfnoise")

B <- 10  

results_list <- list()
iter <- 1

for (EPV in EPV_vals) {
  for (P in P_vals) {
    for (rho in rho_vals) {
      for (bp in beta_patterns) {
        for (b in 1:B) {
          
          # Determine sample sizes 
          N  <- ceiling((EPV * P) / event_frac)
          N_star <- ceiling(5000 / event_frac)   
          
          # Training data 
          train_data <- data_simulation_function(N, P, rho, bp)
          
          # Test data
          test_data  <- data_simulation_function(N_star, P, rho, bp)
          
          # fit model
          fit_model <- model_fitting_function(train_data)
          
          # evaluate model
          metrics <- model_assessment_function(fit_model, test_data)
          
          metrics$model <- "MLE"  
          
          # metadata
          metrics <- metrics %>%
            mutate(
              model = "MLE",
              EPV = EPV,
              P = P,
              rho = rho,
              beta_pattern = bp,
              sim = b
            )
          
          # results
          results_list[[iter]] <- metrics
          iter <- iter + 1
        }
      }
    }
  }
}
results_df_mle <- dplyr::bind_rows(results_list); results_df_mle
write.csv(results_df_mle, mle_path, row.names = FALSE)

```

```{r read mle csv}
# read file

mle_results <- read.csv(mle_path)

```

```{r}

roc_obj <- roc(dat_mle$y, predict(fit_mle, newdata = dat_mle, type="response"))
plot(roc_obj, col="rosybrown", lwd=2)

```

```{r mle calibration}
# MLE calibration plot

cal_df <- dat_mle %>%
  mutate(prob = predict(fit_mle, newdata = dat_mle, type="response"),
         decile = ntile(prob, 10)) %>%
  group_by(decile) %>%
  summarize(obs = mean(y), pred = mean(prob))

ggplot(cal_df, aes(pred, obs)) +
  geom_point(size = 3) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, color = "yellow4", lty = 2) +
  ylim(0,1) + xlim(0,1) +
  ggtitle("Calibration Plot")

```

\hstart

## 2. Ridge Regression

\hstop

```{r ridge regression}
# ridge regression

fit_logistic_ridge <- function(dat) {
  # dat: data.frame with first column y and the rest predictors
  
  X <- as.matrix(dat[ , -1])  # all columns except the first
  y <- dat$y                  # first column is y
  
  # alpha = 0 -> ridge penalty
  cvfit <- cv.glmnet(
    x = X,
    y = y,
    alpha = 0, 
    family = "binomial"
  )
  
  cvfit
}

```

```{r evaluate ridge model}
# ridge-specific evaluate model
# evaluating AUC and brier scores

evaluate_ridge_model <- function(cvfit, dat) {
  X <- as.matrix(dat[ , -1])
  y <- dat$y
  
  # lambda.min from CV (the value that minimizes the error)
  probs <- predict(cvfit, newx = X, type = "response", s = "lambda.min")
  
  auc_val <- pROC::auc(y, probs)
  brier <- mean((y - probs)^2)
  
  tibble::tibble(
    AUC = as.numeric(auc_val), # AUC: area under the ROC curve
    Brier = brier, # brier: measure of the accuracy of probabilistic predictions
    lambda = cvfit$lambda.min
  )
}

```

```{r test run}
# testing the ridge model for one run

set.seed(400)

# simulating one dataset 
dat_test <- simulate_logit(N = 1000, P = 8, rho = 0.5, beta_pattern = "strong")

ridge_fit <- fit_logistic_ridge(dat_test)
ridge_summary <- evaluate_ridge_model(ridge_fit, dat_test)

ridge_summary

```

```{r ridge_mc, eval=FALSE}
# Monte Carlo simulation settings (subset of Van Calster et al. design)

EPV_vals <- c(5, 10, 50)    # subset -> research study uses 7 values
event_frac <- 0.5           # subset -> research study uses 4 fractions
P_vals <- c(4, 8, 12)       # same p-vals as study
rho_vals <- c(0, 0.5)       # same continuous predictors
beta_patterns <- c("equal", "strong", "noise", "halfnoise")   # same beta patterns

B <- 10  # number of Monte Carlo repetitions per condition


# Monte Carlo loop for ridge regression

set.seed(400)

ridge_results <- list()

iter <- 1

for (EPV in EPV_vals) {
  for (P in P_vals) {
    for (rho in rho_vals) {
      for (bp in beta_patterns) {
        for (b in 1:B) {
          
          # compute sample size N from EPV formula:
          # EPV = (N * event_frac) / P => N = EPV * P / event_frac
          N <- ceiling(EPV * P / event_frac)
          
          ## NEW FOR TESTING
          N_star <- ceiling(5000 / event_frac)
          
          # NEW FOR TESTING: SIMULATE TRAINING & TESTING DATA
          df_train <- simulate_logit(N = N, P = P, rho = rho, beta_pattern = bp)
          df_test  <- simulate_logit(N = N_star, P = P, rho = rho, beta_pattern = bp)
        
          # simulate one dataset -> USED ON TRAINING
          dat <- simulate_logit(N = N, P = P, rho = rho, beta_pattern = bp)
        
          # fit ridge model -> NEW fit on df_train
          ridge_fit <- fit_logistic_ridge(df_train)
        
          # evaluate performance on this dataset
          ridge_summary <- evaluate_ridge_model(ridge_fit, df_test)
        
          # store results
          ridge_results[[iter]] <- tibble(
            model = "ridge",
            EPV = EPV,
            sim = b, 
            P = P, 
            rho = rho,
            beta_pattern = bp,
            AUC = ridge_summary$AUC,
            Brier = ridge_summary$Brier,
            lambda = ridge_summary$lambda
          )
          iter <- iter + 1
        }
      
      }
    
    }
  
  }
  
}


ridge_results_df <- bind_rows(ridge_results)
head(ridge_results_df)
write.csv(ridge_results_df, file = ridge_path, row.names = FALSE)

```

```{r read ridge_csv}
# read file

ridge_results <- read.csv(ridge_path)

```

```{r ridge summary stats}

# matching the research paper summary stats

ridge_condition_summary <- ridge_results |>
  group_by(EPV, P, rho, beta_pattern) |>
  summarise(
    n_sims = n(),
    mean_AUC = mean(AUC),
    sd_AUC = sd(AUC),
    mean_Brier = mean(Brier), 
    sd_Brier = sd(Brier),
    mean_lambda = mean(lambda),
    .groups = "drop"
  )

ridge_condition_summary
knitr::kable(ridge_condition_summary, digits = 3)

```

```{r ridge AUC plot}
# AUC plot

ggplot(ridge_condition_summary,
       aes(x = factor(P), y = mean_AUC, group = factor(rho), color = factor(rho))) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ beta_pattern) +
  labs(
    title = "Ridge Regression: Mean AUC by Condition",
    x = "Number of Predictors (P)",
    y = "Mean AUC",
    color = "Correlation (rho)"
  ) +
  geom_ribbon(aes(ymin = mean_AUC - sd_AUC,
                ymax = mean_AUC + sd_AUC,
                fill = factor(rho)),
            alpha = 0.30, color = NA)

```

```{r Brier plot}
# Brier plot

ggplot(ridge_condition_summary,
       aes(x = factor(P), y = mean_Brier, group = factor(rho), color = factor(rho))) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ beta_pattern) +
  labs(
    title = "Ridge Regression: Mean Brier Score by Condition",
    x = "Number of Predictors (P)",
    y = "Mean Brier Score",
    color = "Correlation (rho)"
  ) +
  geom_ribbon(aes(ymin = mean_Brier - sd_Brier,
                ymax = mean_Brier + sd_Brier,
                fill = factor(rho)),
            alpha = 0.15, color = NA)
  


```

\hstart

### Ridge Calibration Plot

\hstop

\pstart

For illustration purposes we show a calibration plot for a SINGLE representative scenario:
$N = 1000, \\P = 8, \\rho = 0.5$, 
beta pattern = $"strong"$.

\pstop

```{r ridge calibration setup}
# calibration plot - how well do the predicted values line up with the actual observed values

set.seed(400)

# simulating a single dataset for calibration
dat_cal <- simulate_logit(
  N = 1000,
  P = 8, 
  rho = 0.5,
  beta_pattern = "strong"
)

# fit ridge logistic regression
ridge_fit_cal <- fit_logistic_ridge(dat_cal)

```

```{r ridge calibration data}

# design matrix for glmnet
X_cal <- as.matrix(dat_cal[ , -1])

# predicted probabilities from the ridge model
ridge_probs <- predict(
  ridge_fit_cal,
  newx = X_cal,
  type = "response", 
  s = "lambda.min"
)

# build calibration dataframe
ridge_cal_df <- dat_cal |>
  mutate(prob = as.numeric(ridge_probs),
         decile = ntile(prob, 10)) |>
  group_by(decile) |>
  summarise(obs = mean(y), pred = mean(prob), .groups = "drop")

ridge_cal_df
  
```

\hstart

## 3. Random Forest

\hstop

```{r}
# Random Forest simulate logit function
# requires factoring

simulate_logit_rf <- function(N, P, rho, beta_pattern) {
  Sigma <- matrix(rho, P, P)
  diag(Sigma) <- 1
  
  X <- mvrnorm(n = N, mu = rep(0, P), Sigma = Sigma)
  
  if (beta_pattern == "equal") {
    beta <- rep(0.5, P)
  } else if (beta_pattern == "strong") {
    beta <- c(1.0, rep(0.2, P-1))
  } else if (beta_pattern == "noise") {
    beta <- c(0, rep(0.3, P-1))
  } else if (beta_pattern == "halfnoise") {
    beta <- c(rep(0, P/2), rep(0.3, P/2))
  } else {
    stop("Unknown beta pattern")
  }
  
  eta <- X %*% beta
  pi  <- 1/(1 + exp(-eta))
  y <- as.factor(rbinom(N, 1, pi))
  
  data.frame(y = y, X)
}

```

```{r random forest}
# random forest fit model

fit_rf_model <- function(dat) {
  rf_spec <- rand_forest(mtry = sqrt(.cols())) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

  rf_fit <- rf_spec |> fit(y ~ ., data = dat)
  rf_fit
}

```

```{r}

fit_rf_model <- function(dat) {
  rf_spec <- rand_forest(mtry = sqrt(.cols())) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

  rf_fit <- rf_spec |> fit(y ~ ., data = dat)
  rf_fit
}

```

```{r}

evaluate_rf_model <- function(fit, dat) {
  probs <- fit |> augment(new_data = dat) |> pull(.pred_1)
  
  auc_val <- auc(as.numeric(dat$y), as.numeric(probs))
  brier <- mean((as.numeric(dat$y) - as.numeric(probs))^2)
  
  tibble(AUC = as.numeric(auc_val), Brier = brier)
}

```

```{r}
  # generate data 
  EPV <- 50
  event_frac <- 0.5
  P <- 12
  rho <- 0.5
  bp <- "equal"
  
  # time model fit
  start <-Sys.time()
  
  # Determine total sample size
  N <- ceiling((EPV * P) / event_frac)
  N_star <- ceiling(5000/event_frac)
  
  # Simulate dataset
  df_train <- simulate_logit_rf(N, P, rho, bp)
  df_test <- simulate_logit_rf(N_star, P, rho, bp)
  
  # fit model
  rf <- fit_rf_model(df_train)
  
  # evaluate model
  results <- evaluate_rf_model(rf, df_test)
  
  end <-  Sys.time()
  end - start
  # about 2 seconds per model * 10 simulations * 72 models = 1,4000 seconds or 24 minutes
  
```

```{r forest_mc, eval=FALSE}
# run for all datasets

set.seed(400)

EPV_vals <- c(5, 10, 50)
event_frac <- 0.5

P_vals <- c(4, 8, 12)
rho_vals <- c(0, 0.5)
beta_patterns <- c("equal", "strong", "noise", "halfnoise")

B <- 10  # number of Monte Carlo repetitions

save_results <- data.frame()

for (EPV in EPV_vals) {
  for (P in P_vals) {
    for (rho in rho_vals) {
      for (bp in beta_patterns) {
        for (b in 1:B) {
          
          # Determine total sample size
          N <- ceiling((EPV * P) / event_frac)
          N_star <- ceiling(5000/event_frac)
          
          # Simulate dataset
          df_train <- simulate_logit_rf(N, P, rho, bp)
          df_test <- simulate_logit_rf(N_star, P, rho, bp)
          
          # fit random forest
          fit <- fit_rf_model(df_train)
          
          # evaluate random forest
          eval_rf <- evaluate_rf_model(fit, df_test)
          
          ### Store combined ###
          combined <- eval_rf %>%
            mutate(
              EPV = EPV,
              P = P,
              rho = rho,
              beta_pattern = bp,
              sim = b
            )
          save_results <- rbind(save_results, combined)
        }
      }
    }
  }
}

save_results$model <- "Random Forest"
save_results$lambda <- NA
write.csv(save_results, rf_path, row.names = FALSE)

```

```{r read forest csv}
# read file

forest_results <- read.csv(rf_path)

```


\hstart

# Comparisons

\hstop

```{r load_mc_results}
mle_df   <- read.csv(mle_path)
ridge_df <- read.csv(ridge_path)
rf_df    <- read.csv(rf_path)

# Standardize model names (so plots don’t look chaotic)
mle_df$model   <- "MLE"
ridge_df$model <- "Ridge"
rf_df$model    <- "Random Forest"

```

\hstart

## Maximum Likelihood vs. Ridge Regression

\hstop

```{r maximum likelihood vs ridge}
# Maximum Likelihood vs Ridge

fit_logistic_ridge <- function(dat) {
  X <- as.matrix(dat[, -1])
  y <- dat$y
  
  cv.glmnet(x = X, y = y, alpha = 0, family = "binomial")
}

evaluate_ridge_model <- function(cvfit, dat) {
  X <- as.matrix(dat[, -1])
  y <- dat$y
  
  probs <- predict(cvfit, newx = X, type = "response", s = "lambda.min")
  
  auc_val <- auc(y, probs)
  brier <- mean((y - probs)^2)
  
  tibble(
    AUC = as.numeric(auc_val),
    Brier = brier,
    lambda = cvfit$lambda.min
  )
}

```

```{r eval=FALSE}
# slow code chunk

set.seed(400)

EPV_vals <- c(5, 10, 50)
event_frac <- 0.5    # FIXED

P_vals <- c(4, 8, 12)
rho_vals <- c(0, 0.5)
beta_patterns <- c("equal", "strong", "noise", "halfnoise")

B <- 10  # number of Monte Carlo repetitions

all_results <- list()
iter <- 1

for (EPV in EPV_vals) {
  for (P in P_vals) {
    for (rho in rho_vals) {
      for (bp in beta_patterns) {
        for (b in 1:B) {
          
          # Determine total sample size
          N <- ceiling((EPV * P) / event_frac)
          
          # Simulate dataset
          dat <- simulate_logit(N, P, rho, bp)
          
          ### MLE ###
          fit_mle <- fit_logistic_mle(dat)
          eval_mle <- evaluate_mle_model(fit_mle, dat)
          eval_mle$model <- "MLE"
          
          ### Ridge ###
          fit_ridge <- fit_logistic_ridge(dat)
          eval_ridge <- evaluate_ridge_model(fit_ridge, dat)
          eval_ridge$model <- "Ridge"
          
          ### Store combined ###
          combined <- bind_rows(eval_mle, eval_ridge) %>%
            mutate(
              EPV = EPV,
              P = P,
              rho = rho,
              beta_pattern = bp,
              sim = b
            )
          
          all_results[[iter]] <- combined
          iter <- iter + 1
        }
      }
    }
  }
}

results_df <- bind_rows(all_results)

```

\pstart

## Main Findings

1. When EPV is small (EPV=5), Ridge performs better overall than MLE, showing:
* higher or similar AUC
* lower Brier scores
* much lower variance
* more realistic estimates when predictors are correlated

2. At EPV=10, both methods perform comparably,
* Although, Ridge remains slightly more stable, especially in the presence of noisy predictors.

3. At EPV=50, MLE and Ridge are almost identical, because shrinkage is no longer needed.

4. Ridge's adaptive penalty $(\lambda)$ adjusts appropriately:
* strong shrinkage at low EPV
* minimal shrinkage at high EPV

\pstop

```{r}
# binding mle & ridge

mle_ridge_results <- bind_rows(mle_df, ridge_df)

```

```{r boxplots mle_ridge}

ggplot(mle_ridge_results, aes(x = model, y = AUC, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_grid(EPV ~ P) +
  labs(title = "AUC Distribution (MLE vs Ridge)", x = "Model", y = "AUC")

ggplot(mle_ridge_results, aes(x = model, y = Brier, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_grid(EPV ~ P) +
  labs(title = "Brier Distribution (MLE vs Ridge)", x = "Model", y = "Brier")

```

```{r heatmaps mle_ridge}

mean_mle_ridge <- mle_ridge_results |>
  group_by(model, EPV, P) |>
  summarise(mean_AUC = mean(AUC), mean_Brier = mean(Brier), .groups = "drop")

ggplot(mean_mle_ridge, aes(x = factor(P), y = factor(EPV), fill = mean_AUC)) +
  geom_tile() +
  facet_wrap(~ model) +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Mean AUC (MLE vs Ridge)", x = "P", y = "EPV", fill = "Mean AUC")

ggplot(mean_mle_ridge, aes(x = factor(P), y = factor(EPV), fill = mean_Brier)) +
  geom_tile() +
  facet_wrap(~ model) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Mean Brier (MLE vs Ridge)", x = "P", y = "EPV", fill = "Mean Brier")
```

\hstart

## Random Forest vs. Maximum Likelihood & Ridge Regression

\hstop

```{r bind all models}

all_results <- bind_rows(mle_df, ridge_df, rf_df)

ridge_extra <- read.csv(ridge_extra_path)
mle_extra   <- read.csv(mle_extra_path)
rf_extra    <- read.csv(rf_extra_path)

```

```{r extra_metrics_plots, out.width="100%"}

ridge_extra$model <- "Ridge"
rf_extra$model <- "Random Forest"
mle_extra$model <- "ML"

combined_metrics <- rbind(ridge_extra, mle_extra)
combined_metrics <- rbind(combined_metrics, rf_extra)

fig2 <- combined_metrics |> pivot_longer(cols = c(rMPSE, MAPE, Brier), 
                                         names_to = "measure",
                                         values_to = "value") |>
  mutate(measure = factor(measure, level = c("rMPSE", "MAPE", "Brier")),
         model = factor(model, level = c("Ridge", "ML", "Random Forest")))

ggplot(fig2) + 
  geom_boxplot(aes(x = model, y = value)) + 
  facet_wrap(measure ~ EPV, scale = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", caption = "predictor selection strategies perform worse")
```

```{r}
fig3 <- combined_metrics  |>
  mutate(model = factor(model, level = c("Ridge", "ML", "Random Forest")),
         P = factor(P),
        "Predictor effects" = factor(beta_pattern, level = c("equal", "strong","noise", "halfnoise")),
        "Dist. Predictors" = factor(rho))|> 
  pivot_longer(cols = c(P, "Predictor effects", "Dist. Predictors"), 
                             names_to = "parameter",
                            values_to = "stage") #|>
  #mutate(parameter = factor(measure, level = c("P", "Dist. Predictors", "Predictor effects")))
ggplot(fig3) + 
  geom_tile(aes(x = stage, y = model, fill = rMPSE))+ 
  scale_fill_distiller(palette = "RdYlGn") +
  facet_wrap( ~ parameter, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplot(fig3) + 
  geom_tile(aes(x = stage, y = model, fill = MAPE))+ 
  scale_fill_distiller(palette = "RdYlGn") +
  facet_wrap( ~ parameter, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


\hstart

# Conclusion

\hstop

\pstart

results:
- improved predictive performance when EPV increased
- little association between Brier and EPV beyond EPV 20
- CIL consistently close to perfect average except for Firth
- CS improved with increasing EPV
- all models except for Ridge showed signs of overfitting (CS above 1)
- impact of shrinkage lessened with increasing EPV
- little variation in Brier and CIL between shrinkage strategies

\pstop

\hstart

# References

\hstop

\pstart

van Smeden M, Moons KG, de Groot JA, et al. Sample size for binary logistic prediction models: Beyond events per variable criteria. Statistical Methods in Medical Research. 2018;28(8):2455-2474. doi:10.1177/0962280218784726

\pstop
